---
title: "example_1_otoliths"
author: "Erin Dillon"
date: "2026-08-23"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Libraries
```{r}
library(rfishbase)
library(tidyverse)
library(ggplot2)
library(patchwork)
library(janitor)
library(hydroGOF)
library(plotrix)
```

## 1. Otolith size - fish size relationships

Fish families in sediment assemblages
```{r}

#all fish families found in Bocas del Toro otolith assemblage dataset
all_families <- c("Acanthuridae","Apogonidae","Atherinidae","Batrachoididae","Bregmacerotidae","Bythitidae","Carangidae","Clupeidae","Engraulidae","Gerreidae","Gobiidae","Haemulidae","Labridae","Mugilidae","Muraenidae","Ophidiidae","Opistognathidae","Pomacentridae","Scaridae","Sciaenidae","Serranidae","Synodontidae")

```

### Reference collection

In-house reference collection data
```{r}

#Load in-house reference collection
oto_size_raw <- read.csv("data/oto_reference_collection_main.csv",header=T)

#Load additional Lutjanus samples from the Atlantic
extra_lutjanus_raw <- read.csv("data/oto_reference_collection_lutjanus.csv",header=T)

```

Clean and combine datasets
```{r}

#Format to be consistent with other data inputs
oto_size <- oto_size_raw %>%
  #use left oto measurement when available, otherwise right
  mutate(otolith_length_mm = ifelse(
    is.na(left.otolith.length.mm)==F, left.otolith.length.mm,right.otolith.length.mm)) %>%
  filter(!Complete_name %in% c("Pomadasys panamensis", "Microspathodon bairdii")) %>% #unusual
  select(Collection_Number,Ocean,Region,Locality,Order,Family,genus,species,Complete_name,total_length_cm,standard_length_cm,otolith_length_mm)


#Lutjanus samples to supplement reference collection
extra_lutjanus <- extra_lutjanus_raw %>% 
  rename(Region=Site,
         otolith_length_mm=Otolith_length_mm) %>% #these are all right otoliths
  mutate(total_length_cm = total_length_mm/10,
         standard_length_cm = standard_length_mm/10) %>% 
  select(Collection_Number,Ocean,Region,Locality,Order,Family,genus,species,Complete_name,total_length_cm,standard_length_cm,otolith_length_mm) 


#Merge datasets and clean header names
oto_ref_col <- rbind(oto_size,extra_lutjanus) %>% 
  filter(!is.na(total_length_cm) & !is.na(otolith_length_mm)) %>% #just keep paired fish-oto measurements
  clean_names()

```

Check available sample sizes
```{r}

#Family-level sample sizes for which both otolith and fish size data are available
oto_ref_col_n <- oto_ref_col %>% 
  group_by(family) %>%
  filter(is.na(otolith_length_mm)==F & is.na(total_length_cm)==F) %>% 
  tally()
oto_ref_col_n


#Add column to clean dataset with family-level sample sizes, for later filtering
oto_ref_col <- oto_ref_col %>% 
  left_join(oto_ref_col_n,by="family")

```

Add AFORO database, source: http://aforo.cmima.csic.es/ (downloaded April 2025)
```{r}

#Load AFORO data
aforo_raw <- read.csv("data/aforo_morphometrics.csv",header=T)

```

Clean AFORO data
```{r}

#Clean and format for consistency with our reference collection data
aforo_clean <- aforo_raw %>% 
  filter(length_type == "TL") %>% #constrain to total length for consistency
  filter(!is.na(fish_length) & !is.na(otolith_length)) %>% #just keep rows with paired fish and oto length data
  mutate(total_length_cm=fish_length/10) %>% #fish length is measured in mm, converting to cm
  rename(otolith_length_mm=otolith_length,
         otolith_width_mm=otolith_width) %>% #oto length & width measured in mm (right)
  mutate(source = "AFORO",.before=family) #adding a source column


#Sample sizes
aforo_n <- aforo_clean %>% 
  group_by(family) %>%
  tally()
aforo_n


#Add column with family-level sample sizes
aforo_clean <- aforo_clean %>% 
  left_join(aforo_n,by="family")

```

Combine size-size dataset
```{r}

#Clean ref col data to be consistent format
oto_ref_col_to_merge <- oto_ref_col %>% 
  mutate(source = "Reference Collection",.before=collection_number) %>% #add source column
  select(-c(order,standard_length_cm)) #remove unneeded columns


#Clean aforo data to match
aforo_to_merge <- aforo_clean %>% 
  rename(collection_number=inst_code) %>% 
  mutate(ocean = NA, #this info was not extracted
         region = NA,
         locality = NA,
         genus = NA,
         species = NA,
         complete_name = NA) %>% 
  select(c("source","collection_number","ocean","region","locality","family","genus","species","complete_name","total_length_cm","otolith_length_mm","n")) #keep the same columns, same order


#Merge datasets
size_size_merged <- rbind(oto_ref_col_to_merge,aforo_to_merge)


#Filter to just include families found in the sediment samples
size_size_merged_corefam <- size_size_merged %>% 
   filter(family %in% all_families)

```

Check available sample sizes in full size-size dataset
```{r}

#sample N for all families
table(size_size_merged$family)
n_distinct(size_size_merged$family)


#sample N for families found in the sediment samples
table(size_size_merged_corefam$source) #n per source
table(size_size_merged_corefam$family) #n per family
length(size_size_merged_corefam$family) #total n
n_distinct(size_size_merged_corefam$family) #n families


size_size_merged_corefam %>% 
  group_by(family,source) %>% 
  tally()


#Sample size comparison
sample_n <- aforo_n %>% 
  full_join(oto_ref_col_n,by="family") %>% 
  rename(aforo_n=n.x,
         ref_col_n=n.y) %>% 
  mutate(aforo_n = replace_na(aforo_n, 0), #replace NA = 0
         ref_col_n = replace_na(ref_col_n, 0)) %>% 
  mutate(total_n = aforo_n+ref_col_n)
sample_n

#Add total_n to combined dataset for filtering
size_size_merged_corefam <- size_size_merged_corefam %>% 
  left_join(sample_n[,c(1,4)],by="family") %>% 
  mutate(family_label = paste0(family," (N=",total_n,")"))

```

### Plot size-size relationships

Fig S2
```{r}

figS2 <- size_size_merged_corefam %>%
  filter(total_n>14) %>% #remove families with n<15
ggplot(aes(y=total_length_cm,x=otolith_length_mm))+
  geom_smooth(aes(color=family_label,fill=family_label),method=lm,se=T,linewidth=0.8,alpha=0.3)+ #family-level model
  geom_smooth(method=lm,color="black",se=T,linewidth=2)+ #overall model
  labs(x="Otolith length (mm)",y="Fish total length (cm)",color="Family",fill="Family")+
  theme_bw()+
   scale_y_continuous(trans='log2')+ 
   scale_x_continuous(trans='log2')+
  theme(axis.text = element_text(size=12),axis.title= element_text(size=16))+
  theme(legend.text = element_text(size=12),legend.title = element_text(size=16))+
  theme(axis.text=element_text(color="black",family="sans"),axis.title= element_text(color="black",family="sans",hjust=0.5))+
  theme(aspect.ratio=1)
figS2

ggsave("outputs/figS2.png",plot=figS2,device="png",height=6)

```

### Power models

#### Otolith length - fish TL
```{r}

#Families to exclude (n<15 data points)
to_exclude <- c("Albulidae","Atherinopsidae","Bregmacerotidae","Hemiramphidae","Labrisomidae","Malacanthidae","Opistognathidae","Paralichthyidae","Priacanthidae")

#List to build family-level models
family_list <- all_families[!(all_families %in% to_exclude)]

```

Aggregate power model (all families with sufficient data)
```{r}

#Exclude families with low n
size_size_merged_corefam2 <- size_size_merged_corefam %>% 
  filter(family %in% family_list)

#Build nls power model
power_fit_all <- nls(total_length_cm~a*otolith_length_mm^b,data=size_size_merged_corefam2, start = list(a=1, b=1))

#View summary and coefficients
summary(power_fit_all)
a_est <- coef(power_fit_all)["a"]
b_est <- coef(power_fit_all)["b"]

#NSE to assess model fit (pseudo r square values)
nse_length_overall <- NSE(sim = fitted(power_fit_all), obs = power_fit_all$m$lhs())

```

Family-level power models
```{r}

#initialize empty list & matrix to store model outputs
allometric_models <- list()
model_coeff <- matrix(NA,nrow=length(family_list),ncol=3)
model_coeff <- as.data.frame(model_coeff)
names(model_coeff) <- c("family","a","b")

#row counter to populate the matrix
counter <- 1

#filter data to include only the specified families with sufficient data
size_size_filtered <- size_size_merged %>% 
   filter(family %in% family_list)
table(size_size_filtered$family) #which families represented, consistent with overall NLS model above
n_distinct(size_size_filtered$family)

#loop through each family to fit a model for each
for (fam in family_list) {
  data_family <- subset(size_size_filtered, family == fam)
  
  #fit an allometric model using nls and store in the allometric_models list
  allometric_model <- nls(total_length_cm~a*otolith_length_mm^b, data = data_family, start = list(a = 1, b = 1))
  allometric_models[[fam]] <- allometric_model
  
  #print the summary of the allometric model for each family
  cat("Allometric model for family:", fam, "\n")
  print(summary(allometric_model))
  cat("\n\n")  #add space between model summaries
  
  #store coefficients in the data frame
  model_coeff[counter,1] <- fam
  model_coeff[counter,2] <- coef(allometric_model)["a"]
  model_coeff[counter,3] <- coef(allometric_model)["b"]
  counter <- counter + 1 #update row counter
}


#check outputs
allometric_models
model_coeff

```

NSE for family-level models to assess fit
```{r}

#initialize an empty vector to store NSE values
nse_values <- numeric(length(allometric_models))
nse_family_length <- numeric(length(allometric_models))

#calculate and store NSE for each model
for (i in seq_along(allometric_models)) {
  family <- names(allometric_models)[i]
  model <- allometric_models[[i]]
  
  #get observed and predicted values
  observed <- model$m$lhs() # observed values
  predicted <- fitted(model) # predicted values
  
  #calculate NSE and print
  nse_values[i] <- NSE(sim = predicted, obs = observed) 
  nse_family_length[i] <- family
  cat("Pseudo R-squared (NSE) for family:", family, "is", nse_values[i], "\n")
}

nse_values

#compile
nse_length <- as.data.frame(cbind(nse_family_length,nse_values)) %>% 
  rename(family = nse_family_length,
         nse_length = nse_values) %>% 
  mutate(nse_length=as.numeric(nse_length)) %>% 
  mutate(nse_length=round(nse_length,3))

```

#### Otolith width - fish TL

Aforo only given data availability
```{r}

aforo_width_corefam <- aforo_clean %>% 
  filter(family %in% family_list) #just families with sufficient data

```

Build aggregate power model
```{r}

#nls power model
power_fit_all_width <- nls(total_length_cm~a*otolith_width_mm^b,data=aforo_width_corefam, start = list(a=1, b=1))

#model summary and coefficients
summary(power_fit_all_width)
a_est_width <- coef(power_fit_all_width)["a"]
b_est_width <- coef(power_fit_all_width)["b"]

#NSE
nse_width_overall <- NSE(sim = fitted(power_fit_all_width), obs = power_fit_all_width$m$lhs())

```

Family-level power models
```{r}

#initialize empty list & matrix to store model outputs
allometric_models_width <- list()
model_coeff_width <- matrix(NA,nrow=length(family_list),ncol=3)
model_coeff_width <- as.data.frame(model_coeff_width)
names(model_coeff_width) <- c("family","a","b")

#row counter to populate the matrix
counter <- 1

#loop through each family to fit a model for each
for (fam in family_list) {
  data_family <- subset(aforo_width_corefam, family == fam)
  
  #fit an allometric model using nls and store in the allometric_models list
  allometric_model_width <- nls(total_length_cm~a*otolith_width_mm^b, data = data_family, start = list(a = 1, b = 1))
  allometric_models_width[[fam]] <- allometric_model_width
  
  #print the summary of the allometric model for each family
  cat("Allometric model for family:", fam, "\n")
  print(summary(allometric_model_width))
  cat("\n\n")  #add space between model summaries
  
  #store coefficients in the data frame
  model_coeff_width[counter,1] <- fam
  model_coeff_width[counter,2] <- coef(allometric_model_width)["a"]
  model_coeff_width[counter,3] <- coef(allometric_model_width)["b"]
  counter <- counter + 1 #update row counter
}

#check outputs
allometric_models_width
model_coeff_width

```

NSE for family-level models to assess fit
```{r}

#initialize an empty vector to store NSE values
nse_values_width <- numeric(length(allometric_models_width))
nse_family_width <- numeric(length(allometric_models_width))

#calculate and store NSE for each model
for (i in seq_along(allometric_models_width)) {
  family <- names(allometric_models_width)[i]
  model <- allometric_models_width[[i]]
  
  #get observed and predicted values
  observed <- model$m$lhs() # Observed values
  predicted <- fitted(model) # Predicted values
  
  #calculate NSE and print
  nse_values_width[i] <- NSE(sim = predicted, obs = observed) 
  nse_family_width[i] <- family
  cat("Pseudo R-squared (NSE) for family:", family, "is", nse_values_width[i], "\n")
}

#compile
nse_width <- as.data.frame(cbind(nse_family_width,nse_values_width)) %>% 
  rename(family = nse_family_width,
         nse_width = nse_values_width) %>% 
  mutate(nse_width=as.numeric(nse_width)) %>% 
  mutate(nse_width=round(nse_width,3))

```
## 2. Bocas fish biomass reconstruction

families with low NSE
```{r}

#remove families with low NSE values = using aggregate model instead
low_NSE <- c("Acanthuridae", "Carangidae", "Clupeidae", "Synodontidae")

low_NSE_width <- c("Acanthuridae", "Batrachoididae", "Carangidae", "Clupeidae", "Gerreidae", "Haemulidae", "Pomacentridae", "Scaridae", "Synodontidae")

```

source functions to predict fish size from otolith size
```{r}

source("functions/oto_fish_pred.R")

```

### Load and clean Caribbean otolith assemblage data

Size data were collected in stages (r1, r2, r3) so require some cleaning and harmonizing
```{r}

#read in raw data, data collection round 1
oto_carib_size_r1_raw <- read.csv("data/carib_oto_meas_r1.csv",header=T)

#cleaning and filter to sites of interest
oto_carib_size_r1_clean <- oto_carib_size_r1_raw %>% 
  clean_names() %>% 
  filter(region=="Bocas") %>% #constrain to Bocas del Toro
  filter(type_age=="Sub Recent") %>% #constrain to modern reefs 
  filter(!locality %in% c('Airport Point','Playa Estrella','SW Bocas Island','Crawl Cay')) #remove, either don't have all otoliths measured or represent diff habitats

```

```{r}

#data collection round 2
oto_carib_size_r2_raw <- read.csv("data/carib_oto_meas_r2.csv",header=T)

#cleaning
oto_carib_r2_to_merge <- oto_carib_size_r2_raw %>%
  clean_names() %>% 
  filter(region=="Bocas") %>% #constrain to Bocas
  filter(type_age=="Sub Recent") %>% #constrain to modern reefs 
  filter(!locality %in% c('Airport Point','Playa Estrella','SW Bocas Island','Crawl Cay')) %>% #remove
  filter(!sample=="AFH12-1-2-3-R") %>% 
  #keep same headers for merging
  select(names(oto_carib_size_r1_clean))

#to track which measurements were added in the combined df
samples_merged  <- oto_carib_size_r1_clean %>% 
  mutate(source="orig_df") %>% 
  select(family,sample,oto_length_mm,source)

#combine and just keep the non-duplicate rows
oto_carib_size_clean_combined <- bind_rows(oto_carib_size_r1_clean,oto_carib_r2_to_merge) %>% 
  distinct() %>% #remove duplicates
  left_join(samples_merged, by=c("family","sample","oto_length_mm")) %>%  #to track what was added
  mutate(oto_width_mm = minor_axis_length/100) %>% #make variable for otolith width
  select(-c(major_axis_length,minor_axis_length)) #remove a couple of extraneous columns

```

```{r}

#data collection round 3
oto_carib_size_r3_raw <- read.csv("data/carib_oto_meas_r3.csv",header=T)

#make 'N/A' values read as actual NAs (currently they are being read as character values)
oto_carib_size_r3_raw$length_um <- oto_carib_size_r3_raw$length_um %>%
  na_if("N/A") %>%      
  as.numeric()

oto_carib_size_r3_raw$width_um <- oto_carib_size_r3_raw$width_um %>%
  na_if("N/A") %>%      
  as.numeric()


#filter and format to match other dfs
oto_carib_clean_r3 <- oto_carib_size_r3_raw %>%
  mutate(region=case_when(
    region=="Bocas del Toro" ~ "Bocas"),
    source="new_2025",
    oto_length_mm=length_um/1000,
    oto_width_mm=width_um/1000,
    type_age=case_when(
      age_group == "modern_bulk" ~ "Sub Recent",
      age_group == "Holocene_bulk" ~ "Holocene")) %>% #to make columns consistent with other df
  rename(locality=site_name,
         oto_length_um=length_um,
         oto_width_um=width_um) %>% 
  mutate(family = case_when( #correcting spelling
    family=="Egraulidae" ~ "Engraulidae",
    TRUE ~ family)) %>% 
  filter(type_age=="Sub Recent") %>% #constrain to modern reefs 
  filter(!locality %in% c('Airport Point','Playa Estrella','SW Bocas Island','Crawl Cay')) %>% #remove
  select(names(oto_carib_size_clean_combined))
  
```

combine datasets
```{r}

#merge datasets
oto_carib_size_clean_combined2 <- rbind(oto_carib_size_clean_combined,oto_carib_clean_r3) %>% 
  mutate(family = case_when( #correcting one more spelling error
    family=="Scianidae" ~ "Sciaenidae",
    TRUE ~ family)) %>% 
  #streamline site names for comparison
  mutate(locality = case_when(
    locality == "Casa Blanca new" ~ "Casa Blanca",
    TRUE ~ locality))%>% 
  #remove a couple families with problematic IDs which were changed during later taxonomic verification
  filter(!family %in% c("Muraenidae","Carangidae","Sphyraenidae","Cynoglossidae","Macrouridae","Mugilidae","Sciaenidae"))

```

clean Caribbean data
```{r}

oto_carib_size_clean <- oto_carib_size_clean_combined2

#sample sizes
nrow(oto_carib_size_clean) #N=1557 otos
length(unique(oto_carib_size_clean$sample)) #42

```

save file
```{r}

write.csv(oto_carib_size_clean,"data_processed/oto_carib_size_clean.csv")

```

### Predict fish length
```{r}

oto_carib_size_length <- oto_carib_size_clean %>%
  #create variable for the predicted fish total lengths
mutate(fish_length_cm = case_when(
  !is.na(oto_length_mm) ~ mapply(predict_allometric,
                                  fam = .data$family,
                                  oto = .data$oto_length_mm),
  is.na(oto_length_mm) ~mapply(predict_allometric_width, #predicts from width if length is not available
                                  fam = .data$family,
                                  oto = .data$oto_width_mm))) %>%
  filter(!is.na(fish_length_cm)) #remove any other broken otoliths for which size could not be reconstructed

#correct taxonomy
oto_carib_size_length <- oto_carib_size_length %>% 
  mutate(family = case_when(
    family=="Scaridae" ~ "Labridae",
    TRUE ~ family))

```

save file
```{r}

write.csv(oto_carib_size_length,"data_processed/oto_carib_size_length.csv")

```

### Predict fish biomass

Compile species list for Caribbean Panama: https://biogeodb.stri.si.edu/caribbean/en/pages
```{r}

#Caribbean species list (SW Caribbean Ecoregion, downloaded 6/11/25)
caribbean_species <- read.csv("data/species_list_caribbean_swecoregion.csv",header=T)

caribbean_species <- caribbean_species %>% 
  mutate(basin="Caribbean")

```

Add higher level taxonomic info from FishBase, using `rfishbase` package
```{r}

#load taxonomic data from fishbase (takes a bit of time to load)
taxa_data <- load_taxa()

#now add family info for each species in our list
caribbean_species <- caribbean_species %>% 
  rename(Species=species_name) %>% 
  left_join(taxa_data[,c(2,3,5)],by="Species") %>% 
  select(Family, Genus, everything())

```

Filter list to families present in sediment assemblages
```{r}

species_list_filtered <- caribbean_species %>% 
  filter(Family %in% c(all_families,"Dorosomatidae")) #adding Dorosomatidae since taxonomy has been updated from Clupeidae

```

Compile length-weight scaling coefficients from FishBase, from power model W=aL^b (units: cm-g)
```{r}

#add scaling at the family level, pull from rfishbase using length_weight() function
lwr_data <- length_weight(species_list_filtered$Species)


#filter for valid non-NA coefficients and TL (total length) values to be consistent with fish length predictions
lwr_data <- subset(lwr_data, !is.na(a) & !is.na(b))
lwr_data <- subset(lwr_data, Type == "TL") #just keep coeff for total length measurements


#check available data
length(unique(lwr_data$Species)) 
length(species_list_filtered$Species)


#merge dfs to keep both demographic info with taxonomic metadata
lwr_data <- lwr_data %>% 
  left_join(species_list_filtered[,c(1:3)],by="Species") %>% #add metadata
  select(Family,Species,Type,Locality,a,b) %>% #keep just relevant data
  distinct() #keep unique rows


#Calculate family-level mean coefficients (a,b) 
family_lwr <- lwr_data %>%
  group_by(Family) %>%
  summarize(
    n_species = n_distinct(Species),
    n_estimates = n(),
    mean_a = mean(a, na.rm = TRUE),
    mean_b = mean(b, na.rm = TRUE),
    .groups = "drop"
  ) %>% 
  rename(family=Family) %>%  #for merging
  mutate(family = ifelse(family == "Dorosomatidae", "Clupeidae", family))  #revert taxonomy so it lines up with the otolith classifications


#add data for Bythitidae, see KULBICKI et al. 2005, "A general approach to length-weight relationships for New Caledonian lagoon fishes" https://horizon.documentation.ird.fr/exl-doc/pleins_textes/divers16-09/010067956.pdf

bythid <- tibble(
  family = "Bythitidae",
  n_species = 1, #Dinematichthys spp.
  n_estimates = 55,
  mean_a = 0.0072,
  mean_b = 3.155)

  
#combine with other family-level data  
family_lwr <- rbind(family_lwr, bythid)

```

Pull sample metadata and group
```{r}

#load data
oto_sample_metadata_raw <- read.csv("data/oto_sample_groups.csv",header=T,na.strings=".")

#extract metadata
oto_sample_metadata <- oto_sample_metadata_raw %>% 
  select(unique,group_name,sed_weight_500_kg,sed_weight_whole_sample_kg,total_oto_count) %>% 
  rename(sample=unique) %>% 
  distinct()

#pooled weights
sample_weight_summed <- oto_sample_metadata %>% 
  group_by(group_name) %>% 
  summarize(sed_weight_group = sum(sed_weight_whole_sample_kg,na.rm=T),
            sed_weight500_group = sum(sed_weight_500_kg,na.rm=T)) 

```

Calculate fish biomass
```{r}

#Use mean length-weight coefficients to predict biomass for each otolith
oto_biomass_merged_pre <- oto_carib_size_length %>%
  left_join(family_lwr, by = "family") %>%  #add coefficients for each family
  left_join(oto_sample_metadata,by="sample") %>%  #add sample metadata
  mutate(group_name = case_when( #fix two samples with a missing group name
    sample=="AT13-10-5" ~ "AT13-10_9",
    TRUE ~ group_name)) %>% 
  mutate(fish_weight_g = mean_a * fish_length_cm^mean_b) #estimate biomass
  

#Calculate per capita biomass and production  
oto_biomass_merged <- oto_biomass_merged_pre %>%
  left_join(sample_weight_summed) %>% 
  group_by(locality,group_name) %>% #calculate per sample group
  summarize(biomass_summed = sum(fish_weight_g,na.rm=T), #calculate total biomass per sample
            n_otoliths = sum(!is.na(fish_weight_g)), #total oto count with data, i.e. number of "individuals"
            per_capita_biomass = biomass_summed/n_otoliths, #mean per-capita biomass
            
            #sample-standardized
            sed_weight_summed = unique(sed_weight_group),
            sed_weight500_summed = unique(sed_weight500_group),
            biomass_kg_summed = biomass_summed/sed_weight_summed, #standardized by sample weight (whole sample)
            biomass500_kg_summed = biomass_summed/sed_weight500_summed) %>%  #standardized by sample weight (500um)
  distinct()


#By family
oto_biomass_merged_fam <- oto_biomass_merged_pre %>% 
  group_by(locality,group_name,family) %>% #calculate per sample group and family
  summarize(biomass_summed_fam = sum(fish_weight_g,na.rm=T), #calculate total biomass per sample/family
            n_otoliths_fam = sum(!is.na(fish_weight_g)), #total oto count with data, i.e. number of "individuals"
            per_capita_biomass_fam = biomass_summed_fam/n_otoliths_fam) %>%  #mean per-capita biomass
  distinct() %>% 
  #add overall biomass estimates for each sample group
  left_join(oto_biomass_merged %>% select(group_name,biomass_summed),by=c("group_name","locality")) %>% 
  mutate(perc_biomass_fam = (biomass_summed_fam/biomass_summed)*100) #% contribution of each family to the overall biomass
oto_biomass_merged_fam

```

Save file
```{r}

write.csv(oto_biomass_merged_pre,"data_processed/oto_biomass_merged_pre.csv")

```

#### Proportional biomass by family
```{r}

#Summarize by family across all samples
perc_biomass_fam_summary <- oto_biomass_merged_fam %>% 
group_by(family) %>% 
  summarize(biomass_mean=mean(perc_biomass_fam, na.rm=T), 
            biomass_sd=sd(perc_biomass_fam, na.rm=T),
            biomass_stderr=std.error(perc_biomass_fam, na.rm=T)) 
perc_biomass_fam_summary

```

Save file
```{r}

write.csv(perc_biomass_fam_summary,"data_processed/perc_biomass_fam_summary.csv")

```

## 3. Rank abundance of each family
```{r}

#Load abundance data
oto_families_raw <- read.csv("data/oto_family.csv",header=T,na.strings=".")


#Filter and clean as above
oto_families <- oto_families_raw %>% 
  filter(type=="modern_bulk") %>%
  filter(!site_name %in% c('Airport Point','Playa Estrella','SW Bocas Island','Crawl Cay')) %>% 
  filter(!unique=="AFH12-1-2-3-R") %>%
  mutate(site_name = case_when(
    site_name == "Casa Blanca new" ~ "Casa Blanca",
    TRUE ~ site_name)) %>% 
  #correct Scaridae taxonomy
  mutate(family = case_when(
    family=="Scaridae" ~ "Labridae",
    TRUE ~ family))

#Add grouping
oto_sample_counts <- oto_sample_metadata_raw %>% 
  select(unique,group_name,total_oto_count) %>%
  distinct() %>% 
  group_by(group_name) %>% 
  mutate(oto_grouped = sum(total_oto_count))
  

#First calculate relative abundance per grouped sample
oto_families_sample_summary <- oto_families %>%
  left_join(oto_sample_counts) %>% 
  group_by(group_name,family) %>% 
  summarize(fam_count_grouped = sum(family_count),
            rel_abund = fam_count_grouped/first(oto_grouped)*100) 
  

#Then take the mean for each family
oto_families_summary <-oto_families_sample_summary %>% 
  group_by(family) %>% 
  summarize(abund_mean = mean(rel_abund,na.rm=T),
            abund_stderr=std.error(rel_abund,na.rm=T))

```

Save file
```{r}

write.csv(oto_families_summary,"data_processed/oto_families_summary.csv")

```
