---
title: "example_1_surveys"
author: "Erin Dillon"
date: "2026-08-23"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Libraries
```{r}
library(rfishprod)
library(tidyverse)
library(rfishbase)
library(plotrix)
```

## 0. Read in data
```{r}

#Rotenone
fish_raw <- read.csv("data/rotenone_surveys.csv",header=T,na.strings=".")

#Reef Life Survey (RLS)
rls_fish_raw <- read.csv("data/panama_fish_rsl_6.10.25.csv",skip=71,header=T)

#Growth data (from `example_1_growth.Rmd`)
growth_data <- read.csv("data_processed/growth_data_4.22.26.csv",header=T) %>% 
  select(-X)

#Traits
species_traits <- read.csv("data/species_traits.csv",header=T,na.strings=".")

```

## 1. Prepare survey data and calculate expected fish abundance

### Rotenone
```{r}

#Convert SL to TL when TL is not available (using FishBase length-length coefficients)
taxa_data <- load_taxa()
length_coeff <- length_length(taxa_data$Species)

length_coeff_filt <- length_coeff %>% 
  filter(Length1 %in% c('TL') & Length2 %in% c('SL')) #filtering to just SL-to-TL conversions


#Take genus-level and family-level mean
length_coeff_genus <- length_coeff_filt %>% 
  separate(Species, into = c("Genus", "Species"), sep = " ", remove = FALSE) %>% 
  group_by(Genus) %>%
  summarize(a_length = mean(a,na.rm=T),
            b_length = mean(b,na.rm=T))


length_coeff_family <- length_coeff_filt %>% 
  left_join(taxa_data[,c(2,3,5)],by="Species") %>% #add metadata
  group_by(Family) %>%
  summarize(a_length = mean(a,na.rm=T),
            b_length = mean(b,na.rm=T))
 
 
#Apply to rotenone dataset
rotenone_fish <- fish_raw %>% 
  left_join(length_coeff_genus) %>% 
  mutate(SL_cm = SL_mm/10) %>% 
  mutate(TL_cm_est = a_length + SL_cm*b_length) %>% 
  #fill in with family data where NAs remain
  left_join(length_coeff_family,by="Family",suffix = c("", ".new")) %>% 
  mutate(TL_cm_est = case_when(
    is.na(TL_cm_est) ~ a_length.new + SL_cm*b_length.new,
    TRUE ~ TL_cm_est)) %>%
  select(-a_length, -b_length,-a_length.new, -b_length.new,-TL_mm,-SL_mm,-SL_cm) %>% #just keep TL_cm
  rename(TL_cm=TL_cm_est)
  

#Calculate sample sizes
nrow(rotenone_fish)
length(unique(rotenone_fish$Group))

rotenone_fish %>% 
  group_by(Site) %>% 
  summarize(n=length(unique((Depth_m))))

rotenone_fish %>% 
  group_by(Group) %>% 
  summarize(n=length(Family))

```

### RLS
```{r}

#Filter
rls_fish <- rls_fish_raw %>% 
  filter(class=="Actinopterygii") %>% #just bony fish
  filter(area=="Bocas del Toro") %>% #just Bocas del Toro
  filter(!site_name %in% c("CK14","Crawl Cay","Popa","Rhino Reef","Salt Creek")) %>%  #removing offshore sites for better comparability
  #updating Scaridae taxonomy
  mutate(family = case_when(
    family=="Scaridae" ~ "Labridae",
    TRUE ~ family))
    

#Make each row a fish and deal with binned size data
rls_fish_expanded <- rls_fish %>% 
  select(site_name,survey_id,block,family, species_name,size_class,total) %>% 
  uncount(weights = total) %>% 
  filter(size_class !=0) %>% #remove fish where size was not estimated
  mutate(TL_cm_mid = case_when( #midpoint of each size class
         size_class <15 ~ size_class + 1.25, #2.5cm groups
         size_class >=15 ~ size_class + 2.5))  #jumps to 5cm groups

#Jitter fish sizes within the size class bins
set.seed(35)
rls_fish_expanded <- rls_fish_expanded %>% 
  #lower/upper limits of size classes
  mutate(bin_lower = case_when(
    TL_cm_mid < 17.5 ~TL_cm_mid - 1.25,
    TL_cm_mid >= 17.5 ~TL_cm_mid - 2.5,
    TRUE ~ NA)) %>% 
  mutate(bin_upper = case_when(
    TL_cm_mid < 17.5 ~TL_cm_mid + 1.25,
    TL_cm_mid >= 17.5 ~TL_cm_mid + 2.5,
    TRUE ~ NA)) %>% 
  mutate(TL_cm = runif(n(), bin_lower, bin_upper))


#Add genus column
rls_fish_expanded <- rls_fish_expanded %>% 
  separate(species_name, into = c("genus", "species"), sep = " ", remove = FALSE)


#Calculate sample sizes
nrow(rls_fish_expanded) 
length(unique(rls_fish_expanded$survey_id)) 
table(rls_fish$site_name)

rls_fish_expanded %>% 
  group_by(survey_id) %>% 
  summarize(n=length(family)) %>% 
  arrange(n)

```


### Calculate expected number of fish

survey area:
rotenone - 3.6m2 per grouped station (=1.8*2)
RLS - 500m2 per survey ID (2 blocks of 250m2)

calculate expected N fish in 100m2 survey, N_exp = N_surv * (A_exp/A_surv)
```{r}

#Rotenone
rotenone_expected <- rotenone_fish %>% 
  group_by(Site,Group) %>% 
  summarize(N_surv=n()) %>% #number of fish per grouped station
  mutate(A_surv = 3.6, #area surveyed
         A_exp = 100, #standardized area
         N_exp = round(N_surv*(A_exp/A_surv)))
  

#RLS
rls_expected <- rls_fish %>% 
  group_by(site_name,survey_id) %>% 
  summarize(N_surv=sum(total,na.rm=T)) %>%  #number of fish per survey
  mutate(A_surv = 500, #area surveyed
         A_exp = 100, #standardized area
         N_exp = round(N_surv*(A_exp/A_surv)))

```

## 2. Resample

Source resampling function
```{r}

source("functions/survey_resampler.R")

```

### Rotenone
```{r}

#Add expected number of fish
rotenone_fish_resample <- rotenone_fish %>% 
  left_join(rotenone_expected %>% select(Group,N_exp)) %>% 
  rename(survey_id=Group) #rename for consistency

#Resample
set.seed(23)
boot_data_rotenone <- resample_surveys(rotenone_fish_resample,n_iter = 10)

```

### RLS
```{r}

#Add expected number of fish
rls_fish_resample <- rls_fish_expanded %>% 
  left_join(rls_expected %>% select(survey_id,N_exp))

#Resample
set.seed(23)
boot_data_rls <- resample_surveys(rls_fish_resample,n_iter = 10)

```

### Combine datasets
```{r}

#Format column names consistently
boot_data_rotenone_merge <- boot_data_rotenone %>% 
  mutate(survey_type="rotenone") %>% 
  mutate(bin_lower=NA) %>%
  select(Site,survey_id,survey_type,Family,Genus,Species,TL_cm,bin_lower,N_exp,iteration) %>% 
  rename(site_name=Site)


boot_data_rls_merge <- boot_data_rls %>%
  mutate(survey_type="RLS") %>%
  select(site_name,survey_id,survey_type,family,genus,species_name,TL_cm,bin_lower,N_exp,iteration) %>% 
  rename(Family=family,
         Genus=genus,
         Species=species_name)


#Merge
fish_combined <- rbind(boot_data_rotenone_merge,boot_data_rls_merge)

```

Save file
```{r}

write.csv(fish_combined,"data_processed/fish_combined.csv")

```


## 3. Calculate production

### Add species growth data and traits
```{r}

#Add growth and trait data
fish_combined_traits <- fish_combined %>%
  left_join(growth_data) %>% 
  mutate(sstmean = 28.8) %>%  #mean SST for Bocas, Isla Colon 4.5 m STRI monitoring mean, downloaded 3.2.26
  mutate(Method = "Otolth") %>% 
  #renaming columns for consistency with rfishprod package inputs
  rename(MaxSizeTL = Lmax, #cm
         a = a_mean,
         b = b_mean) %>% 
  left_join(species_traits)

#Check for NAs
anyNA(fish_combined_traits[, c(11:17)]) #FALSE = no NAs

#cleaned version for production estimates
fish_clean <- fish_combined_traits

```

### Biomass
```{r}

#Estimate body mass per fish
fish_clean <- fish_clean %>% 
  mutate(mass_g = a*TL_cm^b) #fish weight in g

```

### Production 

Predict Kmax
```{r}

#Prepare levels
expected_levels_method <- levels(db$Method)
expected_levels_position <- levels(db$Position)
expected_levels_diet <- levels(db$Diet)

fish_clean$Method <- factor(fish_clean$Method, levels = expected_levels_method)
fish_clean$Position <- factor(fish_clean$Position, levels = expected_levels_position)
fish_clean$Diet <- factor(fish_clean$Diet, levels = expected_levels_diet)


#Formula from Morais and Bellwood (2018) #https://onlinelibrary.wiley.com/doi/pdf/10.1111/faf.12297
fmod <- formula(~ sstmean + MaxSizeTL + Diet + Position + Method)


#Predicting Kmax
fish_pred <- predKmax(fish_clean,
                   dataset = db,
                   fmod = fmod,
                   niter = 100, 
                   return = 'pred')

fish_pred <- fish_pred$pred

```

Cleaning step to check lengths vs Lmax prior to estimating growth
```{r}

#Check observations where fish TL>Lmax
fish_pred$size_check = fish_pred$TL_cm/fish_pred$MaxSizeTL
nrow(fish_pred[fish_pred$size_check>1,])

check_lmax <- fish_pred[fish_pred$size_check>1,] %>% 
  select(Family,Genus,Species,TL_cm,MaxSizeTL,survey_type,size_check) %>% 
  distinct()

#Clean dataset
fish_pred_size_checked <- fish_pred %>% 
  filter(size_check<1.8) %>%   #removing some egregious values
  mutate(TL_cm = case_when( #reduce to lower edge of size class
    size_check > 1 ~ bin_lower,
    TRUE~TL_cm)) %>%
  mutate(TL_cm = case_when( #correcting C. personatus, which is still presenting problems
    Species=="Coryphopterus personatus" & TL_cm>4 ~ 3.75, #bump down one size class
    TRUE~TL_cm)) %>%
  mutate(size_check=TL_cm/MaxSizeTL) 
  
nrow(fish_pred_size_checked[fish_pred_size_checked$size_check>1,]) #re-check

#Filter out the remaining problematic values
fish_pred <- fish_pred_size_checked %>% 
  filter(size_check <= 1)

```

Estimate somatic growth per day
```{r}

fish_pred$soma_gain <- with(fish_pred, somaGain(a = a,
                      b = b,
                      Lmeas = TL_cm,
                      Lmax = MaxSizeTL,
                      Kmax = Kmax))

```

Estimate mortality
```{r}

#Predict instantaneous mortality rate
fish_pred$Md <- with(fish_pred,
                    predM (Lmeas = TL_cm,
                             Lmax = MaxSizeTL,
                             Kmax = Kmax,
                             method = 'Lorenzen'))
                           

#Estimate mortality, using the stochastic approach to assign survival fate to each individual
fish_pred$survive <- applyMstoch(fish_pred$Md)

```

Estimate productivity 
```{r}

#Calculate biomass and growth of the surviving individuals
fish_pred_final <- fish_pred %>%
  filter(survive==TRUE) %>% 
  mutate(biomass_prod_indiv=mass_g+soma_gain)  #fish biomass + somatic growth over m days

```

Save file
```{r}

write.csv(fish_pred_final,"data_processed/fish_pred_final.csv")

```

Productivity by site
```{r}

#Calculate assemblage-level standing biomass and production at each site
fish_productivity <- fish_pred_final %>%
  group_by(site_name,survey_id,iteration) %>% #by survey id
  summarize(area_surveyed = 100, #m2, now standardized across survey types
            biomass_summed = sum(mass_g,na.rm=T)/area_surveyed, #g/m2
            soma_gain_summed = sum(soma_gain,na.rm=T)/area_surveyed, #g/m2/day
            n_fish = sum(!is.na(mass_g)),
            per_cap_biomass = sum(mass_g,na.rm=T)/n_fish, #g/fish
            production_summed = soma_gain_summed, #production, cum sum. G - D, g per m2 per day
            production_kg_ha_day = production_summed * 10, #converting to kg/ha/day
            biomass_t_ha = biomass_summed * 0.01) #converting to t/ha


#Plot biomass by productivity
fish_productivity %>% 
  group_by(site_name) %>% 
  summarize(mean_biomass = mean(biomass_t_ha,na.rm=T),
            mean_production = mean(production_kg_ha_day,na.rm=T)) %>% 
ggplot(aes(x=mean_biomass,y=mean_production))+
  geom_point(aes(fill=site_name),size=4,shape=21,color="gray30")+
  theme_bw()+
  labs(x="Biomass (t/ha)",y="Productivity (kg/ha/day)",color="Site",fill="Site")+
  theme(axis.text = element_text(size=12),axis.title= element_text(size=16))+
  theme(axis.text=element_text(color="black",family="sans"),axis.title= element_text(color="black",family="sans"))+
  theme(aspect.ratio = 1)

```

## 4. Summarize across resampling iterations

By iteration
```{r}

fish_productivity_iteration <- fish_pred_final %>%
  group_by(site_name,survey_id,iteration) %>% 
  summarize(area_surveyed = 100, #standardized to 100m2 per survey iteration
            biomass_summed_iter = sum(mass_g,na.rm=T)/area_surveyed, #g/m2
            soma_gain_summed_iter = sum(soma_gain,na.rm=T)/area_surveyed, #g/m2/day
            n_fish_iter = sum(!is.na(mass_g)),
            standing_biomass_summed_iter = biomass_summed_iter+soma_gain_summed_iter, #cum sum M + G - D, g per m2 per day
            production_summed_iter = soma_gain_summed_iter)

```

By family
```{r}

fish_productivity_fam <- fish_pred_final %>%
  group_by(site_name,survey_id,iteration,Family) %>% 
  summarize(area_surveyed = 100, #standardized to 100m2 per survey iteration
            biomass_summed = sum(mass_g,na.rm=T)/area_surveyed, #g/m2
            soma_gain_summed = sum(soma_gain,na.rm=T)/area_surveyed, #g/m2/day
            n_fish = sum(!is.na(mass_g)),
            per_cap_biomass = sum(mass_g,na.rm=T)/n_fish, #g/fish
            standing_biomass_summed = biomass_summed+soma_gain_summed, #standing biomass, cum sum. M + G - D, g/m2/day
            production_summed = soma_gain_summed, #production, cum sum. G - D, g per m2 per day
            turnover = (production_summed/biomass_summed)*100, #see Morais et al 2020, % per day
            production_kg_ha_day = production_summed * 10, #converting to kg/ha/day
            biomass_t_ha = biomass_summed * 0.01) %>%   #converting to t/ha
  #calculating % contribution of each family
  left_join(fish_productivity_iteration %>% select(site_name,survey_id,iteration,biomass_summed_iter,standing_biomass_summed_iter,production_summed_iter,n_fish_iter)) %>% 
  mutate(perc_biomass_fam = (biomass_summed/biomass_summed_iter)*100,
         perc_biomass_production_fam = (standing_biomass_summed/standing_biomass_summed_iter)*100,
         perc_production_fam = (production_summed/production_summed_iter)*100,
         perc_abundance_fam = (n_fish/n_fish_iter)*100) 
fish_productivity_fam


#summarize across iterations for each survey ID, then across families
fish_productivity_fam_survey <- fish_productivity_fam %>% 
  group_by(survey_id,Family) %>%
  summarize(perc_biomass_production_fam = mean(perc_biomass_production_fam,na.rm=T),
            perc_production_fam = mean(perc_production_fam,na.rm=T),
            perc_abundance_fam = mean(perc_abundance_fam,na.rm=T),
            turnover_fam = mean(turnover,na.rm=T)) 

  
fish_productivity_fam_summary<- fish_productivity_fam_survey %>%  
  group_by(Family) %>% 
  summarize(biomass_production_mean = mean(perc_biomass_production_fam,na.rm=T),
            biomass_production_sd = sd(perc_biomass_production_fam,na.rm=T),
            biomass_production_stderr = std.error(perc_biomass_production_fam,na.rm=T),
            production_mean = mean(perc_production_fam,na.rm=T),
            production_sd = sd(perc_production_fam,na.rm=T),
            production_stderr = std.error(perc_production_fam,na.rm=T),
            abundance_mean = mean(perc_abundance_fam,na.rm=T),
            abundance_sd = sd(perc_abundance_fam,na.rm=T),
            abundance_stderr = std.error(perc_abundance_fam,na.rm=T),
            turnover_mean = mean(turnover_fam,na.rm=T),
            turnover_sd = sd(turnover_fam,na.rm=T),
            turnover_stderr = std.error(turnover_fam,na.rm=T))

```

Save file
```{r}

write.csv(fish_productivity_fam_summary,"data_processed/fish_productivity_fam_summary.csv")

```
